---
title: "Achenbach System of Empirically Based Assessment (ASEBA) T-scores"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tlfb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(ABCDscores)
reactable_theme <- reactable::reactableTheme(
  style = list('html[data-bs-theme="dark"] &' = list(color = "#fff", background = "#282a36")),
  cellStyle = list('html[data-bs-theme="dark"] &' = list(borderColor = "rgba(255, 255, 255, 0.15)")),
  headerStyle = list('html[data-bs-theme="dark"] &' = list(borderColor = "rgba(255, 255, 255, 0.15)")),
  paginationStyle = list('html[data-bs-theme="dark"] &' = list(borderColor = "rgba(255, 255, 255, 0.15)")),
  rowHighlightStyle = list('html[data-bs-theme="dark"] &' = list(background = "rgba(255, 255, 255, 0.04)")),
  pageButtonHoverStyle = list('html[data-bs-theme="dark"] &' = list(background = "rgba(255, 255, 255, 0.08)")),
  pageButtonActiveStyle = list('html[data-bs-theme="dark"] &' = list(background = "rgba(255, 255, 255, 0.1)")),
  searchInputStyle = list('html[data-bs-theme="dark"] &' = list(color = "#fff", background = "#65697d")),
  filterInputStyle = list('html[data-bs-theme="dark"] &' = list(color = "#fff", background = "#65697d")),
  inputStyle = list('html[data-bs-theme="dark"] &' = list(color = "#fff", background = "#65697d"))
)
```

## Introduction

The Achenbach System of Empirically Based Assessment, ASEBA
[@Achenbach2017; @Achenbach2019-wy], is a widely used, 
evidence-based tool designed to assess behavioral, emotional, social, and 
adaptive functioning in individuals aged 1½ to 90+ years. Developed by Thomas 
Achenbach in response to the limitations of early psychiatric classifications 
like the DSM-I, it employs a multi-informant approach to capture behavioral 
patterns across different settings (e.g., home, school) and perspectives 
(e.g., parents, teachers, self-reports).

### T-score

T-scores are standardized scores that indicate how an individual's performance 
compares to an age and sex matched normative sample. These scores are derived by converting raw 
scores into a standardized format with a mean of 50 and a standard deviation 
of 10. To establish these norms, ASEBA collected data from large, representative 
samples across various societies, taking into account factors such as age, 
sex, and cultural background. This extensive normative data allows for 
the comparison of an individual's scores to those of a relevant peer group, 
facilitating the identification of behavioral, emotional, social, and thought 
problems [@Achenbach2019-wy].

### ABCD ASEBA forms

The ABCD Study<sup>®</sup> uses six ASEBA forms: 

- Child Behavior Checklist (CBCL)
- Adult Behavior Checklist (ABCL)
- Youth Self-Report (YSR)
- Adult Self-Report (ASR)
- Brief Problem Monitor (BPM; Youth)
- Brief Problem Monitor (BPM; Teacher)

> **Note:** For the 6.0 release, no T-scores are computed for the ASR form (raw
scores are available in the tabulated data). This may change in future releases.

### ABCD T-scores

The following scales are available in the ABCD data resource:

```{r , echo=FALSE}
scores <- ls("package:ABCDscores") %>%
  {
    .[stringr::str_detect(., "[^s]{2}_tscore$")]
  } |>
  stringr::str_remove("^compute_")

tibble::tibble(
  form = scores |>
    stringr::str_remove("mh_._") |>
    stringr::str_extract("([a-z]+)", 1) |>
    toupper(),
  score = scores,
) |>
  dplyr::mutate(
    form = dplyr::case_when(
      form == "BPM" & stringr::str_detect(score, "_y_") ~ "BPM (Youth)",
      form == "BPM" & stringr::str_detect(score, "_t_") ~ "BPM (Teacher)",
      .default = form
    )
  ) |>
  reactable::reactable(
    striped = TRUE,
    filterable = TRUE,
    columns = list(
      form = reactable::colDef(
        name = "ASEBA form"
      ),
      score = reactable::colDef(
        name = "T-score name"
      )
    ),
    defaultColDef = reactable::colDef(
      minWidth = 120
    ),
    defaultPageSize = 25,
    theme = reactable_theme
  )
```

## Usage

To calculate the T-scores for the ASEBA forms, one must obtain the norming
tables. Since these tables are a proprietary product licensed by ASEBA, they are
not included in the `ABCDscores` package. Therefore, we will demonstrate how to
compute the T-scores using a simulated norming table. To use the code in a
real-world setting, you would need to obtain the norming tables from ASEBA.

### The norming table

The ASEBA norm table needs to be organized as a dataframe with the following
columns:

- `sex`: character, either `"1"` for male or `"2"` for female.
- `age_min`: numeric, the minimum age of the norm group.
- `age_max`: numeric, the maximum age of the norm group.
- `scale_r`: numeric, the raw score.
- `scale_t`: numeric, the T-score.

Here is an example for a simulated norming table:
```{r}
data_norm <- tibble::tibble(
  sex = c("1", "1", "1", "1", "1"),
  age_min = 18,
  age_max = 35,
  scale_r = 0:4,
  scale_t = 20:24
)
data_norm
```

### Compute T-scores

To compute the T-scores, we can use the `compute_<score_name>()` functions. The 
input and output and some important arguments of these functions are 
very similar to other score functions in the `ABCDscores` package. Please refer
to the [Get started](ABCDscores.html) page for an overview of how to use these
functions.

The T-score functions have the following additional arguments:

- `data_norm`: tibble, the norming table (as the one generated above).
- `col_age`: character, specifies the column name in the input data
  that contains the age of the participants. This is required to match the age
  of the participant to the age range in the norming table. For example, 
  for score `mh_p_abcl__afs__frnd_tscore`, the default value of `col_age` is 
  `"mh_p_abcl__cg2__age_001"`. Please refer to the
  [reference pages](../reference/index.html#mental-health-aseba-)
  to see the default value of `col_age` for a given score.
- `col_sex`: character, specifies the column name in the input data that
  contains the participants' age. Similarly, this information is required to
  look up the respective T-score. For example, for score
  `mh_p_abcl__afs__frnd_tscore`, the default value of `col_sex` is
  `"mh_p_abcl__cg2_gi"`. Please refer to the
  [reference pages](../reference/index.html#mental-health-aseba-) to see
  the default value of `col_sex` for a given score. 
  
For data preparation, please also refer to the [Get started](ABCDscores.html)
page. After creating a dataset containing the needed columns, downloading it,
and reading it into a data frame, the data should be ready for the T-score
computation. For example, assuming a data frame containing all needed columns
called `data` and the norming table called `data_norm`, we can execute 

```{r eval=FALSE}
compute_mh_p_abcl__afs__frnd_tscore(data, data_norm)
```

to compute the T-score `mh_p_abcl__afs__frnd_tscore` for the ABCL.

### Examples

As we cannot provide the actual norming tables as part of this package, we
cannot run the code above. Here we demonstrate how to compute T-scores using a
simulated norming table and a simulated data table using the lower level
function `ss_tscore()`.

All specific `compute...()` T-score functions use the lower level function
`ss_tscore()`. The only difference between
scale-specific T-score functions are default values, such as `col_age` and
`col_sex`.

Create a simulated data table:

```{r}
data <- tibble::tibble(
  var1 = c(0, 1, NA, 1, 2, 3),
  var2 = c(1, 2, 1, 2, 5, 3),
  age = c(18, 20, 25, 99, 35, 30),
  sex = c("1", "1", "1", "1", "1", "2")
)
data
```

Compute T-scores using the simulated norm and data tables:

```{r}
ss_tscore(
  data = data,
  data_norm = data_norm,
  col_age = "age", # specify the age column
  col_sex = "sex", # specify the sex column
  vars = c("var1", "var2") # specify individual items summarized in the score
)
```

If `ss_tscore()` cannot find a match in the norming table, it will return `NA`
for the T-score. For example, in the above example, the scores for the
following rows are `NA`:

- *row 4:* because the age is 99, which is not in the age range [18, 35].
- *row 5:* because the raw score is 5, which makes the raw score
  sum 7, which is not in the raw score range [0, 4].
- *row 6:* because the sex is "2", which is not included in the 
  simulated norming table.

#### Customize T-score computation

We can use the `max_na` and `exclude` arguments to modify certain aspects of the
T-score computation.

- The `max_na` argument specifies the maximum number of missing values that are
allowed to still compute the T-score. For example, we can set `max_na = 0` to
allow no missing values in the summarized items (the default for
`ss_tscore()` is `NULL`, i.e., no limit, but the specific functions for the
ASEBA T-scores impose a limit). If we rerun the code above with this change,
`NA` is returned for row 3 because `var1` is `NA` in that row.
    ```{r}
    ss_tscore(
      data = data,
      data_norm = data_norm,
      col_age = "age",
      col_sex = "sex",
      vars = c("var1", "var2"),
      max_na = 0 # specify maximum number of missing values
    )
    ```
- The `exclude` argument excludes certain values, i.e, converts them to `NA`
before the score computation. For example, we can set `exclude = "0"` to exclude
any `"0"` values from the computation. If we rerun the code above with this
change, `NA` is returned for row 1 because `var1` is `"0"` in that row.
    ```{r}
    ss_tscore(
      data = data,
      data_norm = data_norm,
      col_age = "age",
      col_sex = "sex",
      vars = c("var1", "var2"),
      max_na = 0,
      exclude = c("0") # specify values to exclude from the computation
    )
    ```

## References
