---
title: "Family History (FAMHX) summary scores"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{famhx}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(ABCDscores)

reactable_theme <- reactable::reactableTheme(
  style = list('html[data-bs-theme="dark"] &' = list(color = "#fff", background = "#282a36")),
  cellStyle = list('html[data-bs-theme="dark"] &' = list(borderColor = "rgba(255, 255, 255, 0.15)")),
  headerStyle = list('html[data-bs-theme="dark"] &' = list(borderColor = "rgba(255, 255, 255, 0.15)")),
  paginationStyle = list('html[data-bs-theme="dark"] &' = list(borderColor = "rgba(255, 255, 255, 0.15)")),
  rowHighlightStyle = list('html[data-bs-theme="dark"] &' = list(background = "rgba(255, 255, 255, 0.04)")),
  pageButtonHoverStyle = list('html[data-bs-theme="dark"] &' = list(background = "rgba(255, 255, 255, 0.08)")),
  pageButtonActiveStyle = list('html[data-bs-theme="dark"] &' = list(background = "rgba(255, 255, 255, 0.1)")),
  searchInputStyle = list('html[data-bs-theme="dark"] &' = list(color = "#fff", background = "#65697d")),
  filterInputStyle = list('html[data-bs-theme="dark"] &' = list(color = "#fff", background = "#65697d")),
  inputStyle = list('html[data-bs-theme="dark"] &' = list(color = "#fff", background = "#65697d"))
)
```


## Introduction

The Family History (FAMHX) survey is a parent survey used to collect information 
about endorsement of substance use and mental health history of the 
participant's family members.

### ❗Data requirements

> **Where to get data?**
>
> The FAMHX item-level data is available to authorized users of the
> [NBDC Data Hub](https://nbdc-datahub.org) as part of the tabulated data resource.
> Please refer to the [Get started](ABCDscores.html) page for general information
> on how to assemble and download a dataset.

For the computation of the FAMHX summary scores, one needs to have access to the 
`mh_p_famhx` table.

## FAMHX summary score functions

We compute two types of FAMHX summary scores (format: `mh_p_famhx_xxxxx`):

1. Summarize the endorsements from one family member (input field) \
  \
  We developed one basic function, `compute_famhx_endorsement()` that can be used 
  to compute all FAMHX summary scores by changing the value of the `vars_matches` 
  parameter, which sets the input variable used to compute the 
  summary score, the name of which is set by changing the value of the `name` 
  parameter. \
  \
  Note: The single input field maps to multiple columns / fields in the 
  release data because each option is provided as its own column. \
  \
2. Summarize the endorsements from multiple family members (input fields) \
  \
  For this set of scores, we first use the `compute_famhx_endorsement()` to 
  compute the endorsement indicators for each input variable and then
  summarize over the multiple endorsements to compute a score. These FAMHX
  scores are listed below.


### One family member (input field)

The following table gives an overview of the FAMHX summary scores that 
are computed from one family member (input field), the associated name of the 
input field and the function call that produces the score:

```{r, echo=FALSE}
reactable::reactable(
  ABCDscores::famhx_config |>
    select(name, input_vars, call),
  filterable = TRUE,
  striped = TRUE,
  columns = list(
    name = reactable::colDef(
      name = "Score name",
      minWidth = 170
    ),
    input_vars = reactable::colDef(
      name = "Input Field (Family Member)",
      minWidth = 170
    ),
    call = reactable::colDef(
      name = "Function call",
      minWidth = 300,
      filterable = FALSE
    )
  ),
  resizable = TRUE,
  theme = reactable_theme,
  style = list(fontSize = "12px")
)
```

The primary computation in the `compute_famhx_endorsement()` is the following,
where,

> `data` = either the entire `mh_p_famhx` table or a subset containing
all the necessary fields as shown above. \
> \
> `combine_checkboxes()` = function that combines the multiple columns for 
an input field, each representing one option, into a single column

```{r, echo=TRUE, eval=FALSE}
data_ss <- data |>
  combine_checkboxes(
    var_basename = var_matches,
    name = "tmp_values"
  ) |>
  transmute(
    !!name := case_when(
      purrr::map_int(tmp_values, length) == 0 ~ NA,
      purrr::map_lgl(tmp_values, ~ 777 %in% .x) ~ NA,
      purrr::map_lgl(tmp_values, ~ 888 %in% .x) ~ NA,
      purrr::map_lgl(tmp_values, ~ 999 %in% .x) ~ NA,
      purrr::map_lgl(tmp_values, ~ 0 %in% .x) ~ 0,
      .default = 1
    ) |>
      as.integer()
  )
```

The classification logic applied to the combined, single field can be read as:

- If no response was provided, or the responses included `777`, `888`, or `999`,
the indicator is set to `NA`
- If the responses included the value `0`, the indicator is set to `0`
- For all other cases, the indicator is set to `1`


### Multiple family members (input fields)

For some of the FAMHX scores (shown below), multiple family members (input 
fields) were summarized to create one endorsement summary score.

```{r, echo=FALSE}
score_funcs <- c(
  "compute_mh_p_famhx__alc__moth__fath_indicator",
  "compute_mh_p_famhx__drg__moth__fath_indicator",
  "compute_mh_p_famhx__dep__moth__fath_indicator",
  "compute_mh_p_famhx__mania__moth__fath_indicator",
  "compute_mh_p_famhx__halluc__moth__fath_indicator",
  "compute_mh_p_famhx__troub__moth__fath_indicator",
  "compute_mh_p_famhx__nerve__moth__fath_indicator",
  "compute_mh_p_famhx__doc__moth__fath_indicator",
  "compute_mh_p_famhx__hosp__moth__fath_indicator",
  "compute_mh_p_famhx__suic__moth__fath_indicator",
  "compute_mh_p_famhx__alc__moth__fath_score",
  "compute_mh_p_famhx__drg__moth__fath_score",
  "compute_mh_p_famhx__dep__moth__fath_score",
  "compute_mh_p_famhx__mania__moth__fath_score",
  "compute_mh_p_famhx__halluc__moth__fath_score",
  "compute_mh_p_famhx__troub__moth__fath_score",
  "compute_mh_p_famhx__nerve__moth__fath_score",
  "compute_mh_p_famhx__doc__moth__fath_score",
  "compute_mh_p_famhx__hosp__moth__fath_score",
  "compute_mh_p_famhx__suic__moth__fath_score"
)

scores_map_tbl <- tibble(
  name = stringr::str_replace_all(score_funcs, "compute_", ""),
  input_vars = stringr::str_replace_all(score_funcs, "compute_", "vars_"),
  func = paste0(
    score_funcs, "()"
  )
) |>
  mutate(
    input_vars = purrr::map_chr(
      input_vars,
      ~ get(.x, envir = asNamespace("ABCDscores")) |>
        paste(collapse = ", ")
    )
  )

reactable::reactable(
  scores_map_tbl,
  filterable = TRUE,
  striped = TRUE,
  columns = list(
    name = reactable::colDef(
      name = "Score name",
      minWidth = 240
    ),
    input_vars = reactable::colDef(
      name = "Input Fields (Family Members)",
      minWidth = 190
    ),
    func = reactable::colDef(
      name = "Name of function",
      minWidth = 220,
      filterable = FALSE
    )
  ),
  resizable = TRUE,
  theme = reactable_theme,
  style = list(fontSize = "12px")
)
```


#### Example 

For example, here are the primary computation steps in the 
`compute_mh_p_famhx__drg__moth__fath_indicator()` function that computes
the `mh_p_famhx__drg__moth__fath_indicator` score, which aggregates
over two individual indicators, `mh_p_famhx__drg__fath_indicator` and 
`mh_p_famhx__drg__moth_indicator`.

```{r eval=FALSE, echo=TRUE}
# Subset the famhx_config file to obtain the
# metadata corresponding to the relevant input fields
config_sub <- ABCDscores::famhx_config |>
  filter(
    input_vars %in% c(
      "mh_p_famhx__drg__fath_001",
      "mh_p_famhx__drg__moth_001"
    )
  )

# Compute the `mh_p_famhx__drg__moth__fath_indicator` score
data_ss <- purrr::map_dfc(
  ## Step 1: compute indicators corresponding to each input field
  config_sub$call,
  ~ eval(parse(text = .x))
) |>
  ## Step 2: aggregate the individual indicators
  transmute(
    !!name := case_when(
      mh_p_famhx__drg__fath_indicator == 1 |
        mh_p_famhx__drg__moth_indicator == 1 ~ 1,
      mh_p_famhx__drg__fath_indicator == 0 &
        mh_p_famhx__drg__moth_indicator == 0 ~ 0,
      .default = NA
    ) |>
      as.integer()
  )
```


The aggregation logic applied above can be read as:

- If either `mh_p_famhx__drg__fath_indicator == 1` or 
  `mh_p_famhx__drg__moth_indicator == 1`, set indicator to `1`.
- If both `mh_p_famhx__drg__fath_indicator` and `mh_p_famhx__drg__moth_indicator`
  are 0, set indicator to `0`.
- For all other cases, the indicator is set to `NA`
