---
title: "Timeline Followback (TLFB) summary scores"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Timeline Followback (TLFB) summary scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(ABCDscores)
trans_uni <- rlang::chr_unserialise_unicode
reactable_theme <- reactable::reactableTheme(
  style = list('html[data-bs-theme="dark"] &' = list(color = "#fff", background = "#282a36")),
  cellStyle = list('html[data-bs-theme="dark"] &' = list(borderColor = "rgba(255, 255, 255, 0.15)")),
  headerStyle = list('html[data-bs-theme="dark"] &' = list(borderColor = "rgba(255, 255, 255, 0.15)")),
  paginationStyle = list('html[data-bs-theme="dark"] &' = list(borderColor = "rgba(255, 255, 255, 0.15)")),
  rowHighlightStyle = list('html[data-bs-theme="dark"] &' = list(background = "rgba(255, 255, 255, 0.04)")),
  pageButtonHoverStyle = list('html[data-bs-theme="dark"] &' = list(background = "rgba(255, 255, 255, 0.08)")),
  pageButtonActiveStyle = list('html[data-bs-theme="dark"] &' = list(background = "rgba(255, 255, 255, 0.1)")),
  searchInputStyle = list('html[data-bs-theme="dark"] &' = list(color = "#fff", background = "#65697d")),
  filterInputStyle = list('html[data-bs-theme="dark"] &' = list(color = "#fff", background = "#65697d")),
  inputStyle = list('html[data-bs-theme="dark"] &' = list(color = "#fff", background = "#65697d"))
)
```

## Introduction

The Timeline Followback (TLFB) [@Sobell1996] is a semi-structured interview method used to assess specific and quantified estimates of substance use over a specified period, approxiamtely the past 12 months in ABCD. It involves participants recalling their substance use while being guided by a Research Assistant (RA). The RA fills out a TLFB calendar, documenting all reports of substance use and helping participants ground their responses to specific dates and events, ensuring accurate reporting of substance use patterns.

## TLFB raw data

The TLFB raw data is available to authorized users of the [NBDC Data Hub](https://nbdc-datahub.org). It combines information from the logfiles of the old and new versions of the TLFB application used in the ABCD assessments (For additional information, please see documentation webpage). The raw data is a long-form data frame with one row per reported substance use instance and the following structure:

```{r}
glimpse(data_tlfb_sample)
```

```{r, echo=FALSE}
data_tlfb_sample |>
  reactable::reactable(
    striped = TRUE,
    defaultColDef = reactable::colDef(
      minWidth = 120
    ),
    columns = list(
      sex = reactable::colDef(
        minWidth = 50
      )
    ),
    theme = reactable_theme
  )
```

The TLFB interview records usage for the following substances:

```{r, echo=FALSE, results='asis'}
paste0("\"", ABCDscores:::tlfb_substances, "\"") |> md_bullet(2, italic = TRUE)
```

where _"Marijuana (all forms)"_ and _"Nicotine (all forms)"_ are aggregating over the different methods of consumption.

## TLFB summary score functions

The ABCD Study's tabulated data resource includes summary scores for the TLFB interview. While the summary scores for other domains typically define one function per summary score, the TLFB summary scores are computed using a set of basic functions, with user-defined parameters. This allows for a more flexible and modular approach to computing the TLFB summary scores, e.g., computing the same score for different substances or time periods. This section describes the basic functions and how to use them to compute the ABCD summary scores using the `ABCDscores` R package. It also describes how users can use these functions to compute TLFB summary scores that are not reported in the ABCD data.

The basic functions are:

- `compute_tlfb_abst()`: Compute abstinence from substance use in days.
- `compute_tlfb_dt()`: Compute first/last day of substance use.
- `compute_tlfb_maxdose()`: Compute maximum dose of substance use.
- `compute_tlfb_mean()`: Compute mean dose of substance use.
- `compute_tlfb_totdose()`: Compute total dose of substance use.
- `compute_tlfb_ud()`: Compute number of substance use days.

The functions take the following parameters:

- `substance` One or several substance(s) to compute the score for.
- `period`: The period (detailed, estimated, or both) for which the score is computed for.
- `days`: Number of days before the TLFB interview to consider.
- `wknd`: Whether the score should be computed only for weekends days, only for week days, or all days
- `co_use`: Only computes score for days on which the substance was taken with specified other substance(s).
- `binge`: Allows to specify binge thresholds to apply, either general or by sex.
- `position`: Whether to consider the first or last substance use

Not all parameters are used by all functions and some parameters are mutally exclusive (see the function documentation for more details). The following gives an overview of which function can use which parameter:

```{r, echo=FALSE}
tribble(
  ~`function`, ~substance, ~period, ~days, ~wknd, ~co_use, ~binge, ~position,
  "compute_tlfb_abst()", trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), NA, NA, trans_uni("<U+2713>"), NA,
  "compute_tlfb_dt()", trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), NA, trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"),
  "compute_tlfb_maxdose()", trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), NA,
  "compute_tlfb_mean()", trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), NA,
  "compute_tlfb_totdose()", trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), NA,
  "compute_tlfb_ud()", trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), trans_uni("<U+2713>"), NA
) |>
  reactable::reactable(
    striped = TRUE,
    defaultColDef = reactable::colDef(
      minWidth = 70
    ),
    columns = list(
      `function` = reactable::colDef(
        minWidth = 170
      ),
      substance = reactable::colDef(
        minWidth = 100
      )
    ),
    theme = reactable_theme
  )
```

### ABCD summary scores

The following table gives an overview of how the basic functions are used to compute the 905 summary scores that are reported in the ABCD data resource (you can use the filter boxes on top of the columns to filter the table):

```{r, echo=FALSE}
tlfb_config |>
  select(
    name,
    `function call` = call
  ) |>
  reactable::reactable(
    filterable = TRUE,
    striped = TRUE,
    # defaultColDef = reactable::colDef(
    #   minWidth = 70
    # ),
    columns = list(
      name = reactable::colDef(
        minWidth = 150
      ),
      `function call` = reactable::colDef(
        minWidth = 200
      )
    ),
    theme = reactable_theme
  )
```

Note that the function calls expect the TLFB raw data to be read into a data frame `data_tlfb` in the global environment. For example, to compute the summary scores, one would run:

```{r, eval=FALSE}
# read raw data
data_tlfb <- arrow::read_parquet("tlfb_raw.parquet")

# compute summary score `su_y_tlfb__alc__1mo__wknd_totdose`
compute_tlfb_totdose(
  data = data_tlfb,
  name = "su_y_tlfb__alc__1mo__wknd_totdose",
  substance = "Alcohol",
  period = NULL,
  days = 30,
  wknd = TRUE,
  co_use = NULL,
  binge = NULL
)

# compute summary score `su_y_tlfb__alc__1mo__wknd_ud`
compute_tlfb_ud(
  data = data_tlfb,
  name = "su_y_tlfb__alc__1mo__wknd_ud",
  substance = "Alcohol",
  period = NULL,
  days = 30,
  wknd = TRUE,
  co_use = NULL,
  binge = NULL
)

# compute all summary scores in the ABCD data resource
data_tlfb_ss <- purrr::map(
  tlfb_config$call,
  ~ eval(parse(text = .x))
) |>
  purrr::reduce(
    full_join,
    by = join_by(
      participant_id,
      session_id
    )
  )
```

### Custom summary scores

The basic functions can also be used to compute custom summary scores that are not reported in the ABCD data resource.

#### Simple examples

*Compute participants' use days for all substances over all events*

```{r, eval=FALSE}
compute_tlfb_ud(
  data = data_tlfb,
  name = "ud_all_subst",
  period = "detailed"
) |>
  summarize(
    ud_all_subst_all_events = sum(ud_all_subst),
    .by = participant_id
  ) |>
  arrange(
    desc(ud_all_subst_all_events)
  )
```

*Compute participants' marijuana or nicotine use days on week days*

```{r, eval=FALSE}
compute_tlfb_ud(
  data = data_tlfb,
  name = "ud_mj_nic_wkdays",
  substance = c("Marijuana (all forms)", "Nicotine (all forms)"),
  period = "detailed",
  wknd = FALSE
)

# ...
```

#### More complex examples

*Compute age of first use for marijuana*

```{r, eval=FALSE}
#' Add age of use
#'
#' @description
#' Adds an `age_use` column to the TLFB raw data based on the anonymized date of
#' birth and the date of use. Age is computed in years (with decimals) using
#' an exact method taking into account leap years (using functions from the
#' `lubridate` package).
#'
#' @param data tibble. The TLFB raw data.
#' @param data_dob tibble. A data frame containing at least two columns, the
#'   participant ID and the anonymized date of birth.
#' @param col_dob character. The name of the column in `data_dob` that has the
#'   anonymized date of birth.
#'
#' @return tibble. The TLFB raw data with an additional column `age_use`.
add_age_use <- function(data, data_dob, col_dob) {
  data_dob <- data_dob |>
    select(
      participant_id,
      dob = {{ col_dob }}
    )

  data |>
    left_join(
      data_dob,
      by = "participant_id"
    ) |>
    mutate(
      age_use = lubridate::interval(dob, dt_use) / lubridate::years(1)
    ) |>
    select(
      -dob
    )
}

# expects that a data frame `data_dob` with a column `dob_anonymized`
# to be available in the global environment; uses the `add_age_use()` function
# to add a column with the age of the participant at each reported use and
# finds the first use of marijuana for each participant
data_tlfb |>
  compute_tlfb_dt(
    name      = "dt_use",
    substance = "Marijuana (all forms)",
    period    = "detailed",
    position  = "first"
  ) |>
  summarize(
    dt_use = min(dt_use),
    .by = participant_id
  ) |>
  add_age_use(data_dob, "dob_anonymized") |>
  select(
    participant_id,
    mj_age_first_use_tlfb = age_use
  )
```

*Compute the longest streak of consecutive weeks with alcohol binge drinking*

```{r, eval=FALSE}
#' Add week sequence
#'
#' @description
#' Adds a `wk_seq` column to the TLFB raw data. The week sequence is computed
#' based on the minimum and maximum dates of use in the data, with week 1 being
#' the one that contains the minimum date of use and the maximum week number
#' being the one that contains the maximum date of use. That is, the week
#' sequence contains consecutive week numbers from the earliest to the latest
#' possible date and works across all years contained in the data. Joining this
#' week sequence to the data allows for the computation of weekly summaries and
#' streaks of consecutive use weeks.
#'
#' @details
#' The week sequence can be computed using either the `lubridate::isoweek()` or
#' the `lubridate::epiweek()` command. The `lubridate::isoweek()` command uses
#' the European standard for week numbering, where Monday is the first day and
#' Sunday is the last day of the week. The `lubridate::epiweek()`command uses
#' the US standard where Sunday is the first day of the week and Saturday is the
#' last day. The default is to use the `lubridate::isoweek()` command as it does
#' not break up a weekend into two different weeks.
#'
#' @param data tibble. The TLFB raw data.
#' @param standard character. The standard to use for week numbering. One of
#'   `"iso"` or `"epi"` (Default: `"iso"`; see details).
#'
#' @return tibble. The TLFB raw data with an additional column `wk_seq`.
add_wk_seq <- function(data, standard = c("iso", "epi")) {
  standard <- match.arg(standard)

  week_command <- switch(standard,
    iso = lubridate::isoweek,
    epi = lubridate::epiweek
  )

  data_weeks <- tibble(
    dt = seq(
      min(data$dt_use),
      max(data$dt_use),
      by = 1
    )
  ) |>
    mutate(
      yr_use = lubridate::year(dt),
      wk_use = week_command(dt)
    ) |>
    distinct(
      yr_use,
      wk_use
    ) |>
    arrange(
      yr_use,
      wk_use
    ) |>
    mutate(
      wk_seq = row_number()
    )

  data |>
    mutate(
      yr_use = lubridate::year(dt_use),
      wk_use = week_command(dt_use)
    ) |>
    left_join(
      data_weeks,
      by = join_by(yr_use, wk_use)
    ) |>
    select(
      -yr_use,
      -wk_use
    )
}

data_tlfb |>
  ABCDscores::filter_tlfb(
    substance = "Alcohol",
    binge = list("M" = 5, "F" = 4),
    period = "detailed"
  ) |>
  add_wk_seq(standard = "iso") |>
  summarize(
    alc_binge = n_distinct(dt_use),
    .by = c(participant_id, session_id, wk_seq)
  ) |>
  arrange(
    participant_id,
    session_id,
    wk_seq
  ) |>
  mutate(
    streak = cumsum(c(TRUE, diff(wk_seq) != 1)),
    .by = c(participant_id, session_id)
  ) |>
  mutate(
    n_wks = row_number(),
    .by = c(participant_id, session_id, streak)
  ) |>
  summarize(
    alc_wks_binge = max(n_wks),
    .by = c(participant_id)
  ) |>
  arrange(
    desc(alc_wks_binge)
  )
```
